The Effect of Elections and Prime Ministers on Discussion in the Australian Federal Parliament (1901--2018)

### 2018-12-15 outputs/docs/events/paper.Rmd
SECOND DRAFT DONE

The first panel of Figure \@ref(fig:threeexamples) is the second Menzies premiership. The dashed lines show the elections. While the results here abstract from differences based on topic mix, the shape of the distribution seems to be reasonably similar. Given 80 topics, it may be difficult to see differences in the topic mix manually, and so we would like for the analysis model to be able to suggest whether the mix is changing. The second panel shows the 1983 election, and the subsequent change from Fraser to Hawke on 11 March 1983, identified by a dotted line on that date. The distribution of topics seems less bunched after the change, but it is difficult to tell how different it is. We would like the analysis model to be able to distinguish between the two if they seem different. The third panel shows the first Rudd premiership. The important aspect to note is that the period after the 2007 election and the Rudd premiership contain essentially the same Hansard dates. The analysis model needs to be able to deal with this situation.

@@

governments_results %>%
  kable(booktabs = T, caption = "Prime ministers that were significantly different to their predecessor") %>%
  kable_styling(font_size = 10) %>%
  footnote(
    general = "The significance of a prime minister is determined by whether at least one topic was significantly different during this premiership, compared with the previous one.",
    footnote_as_chunk = T,
    threeparttable = TRUE)

@@

Of the 36 prime ministers over this period, we find that 20 of them are significantly different to the prime minister that preceded them. However, four of these results -- the significance of the Page premiership, the first Menzies premiership, the Chifley premiership and the Abbott premiership -- are likely due to the short length of either the prime minister or the predecessor and should be ignored, leaving only 16. The earliest of these are the first Fisher premiership, the Cook premiership and the third Fisher premiership, which were different to the second Deakin premiership, the second Fisher Premiership, and the Cook Premiership, respectively. To a certain extent this is likely due to changes due to World War I.

The Second Menzies Premiership, begining in 1949, is the next government that is significantly different to its predecessor. The other governments that are different are concentrated in the second half of our sample, with three of them being in the past twenty years. Similarly, of the 45 general elections that have been held we find that 14 of them define periods that were significantly different to their immediate predecessor. 1974, 1980, 1990, 1998, 2004, and 2007 stand out as elections where the government did not change, but the model suggests there was considerable change in the topics discussed in parliament.

The Second Menzies Premiership was associated with a variety of changes compared with the preceding Chifley Government. The Chifley Premiership had governed through the end of World War II and the difficult economic times that followed. There was also a large increase in the number of seats in the House of Representatives at the 1949 election. Many new politicians entered parliament, and this changed representation may also have been partly to do with the changed distribution of discussion topics, although further investigation of this is left for future work. The sixteen-year length of the Second Menzies Premiership, and better economic conditions over this time make it understandable why parliamentary discussion would have been different. There were six elections within the Second Menzies Premiership Three toward the middle of that government were associated with significant changes in the topics discussed.

The Menzies Premiership was succeeded by the Holt Premiership in January 1966. This is an example where there was a change in prime minister without an election, as the next election only happened in November 1966. We find that the Holt Premiership is different to the Menzies Premiership In Figures \@ref(fig:muelectionall) we compare the topics during the final term of the Menzies Premiership with the topics of the Holt Premiership

@@

The Whitlam Premiership is especially interesting as we find a difference in the topics after it was first elected in December 1972, compared with its second election win in May 1974. Figure \@ref(fig:muelectionall) compares the topics that are significant in the first Whitlam term and then compares them to those in the second Whitlam Premiership.

The Howard Premiership is also interesting because of the significant differences between elections. For instance, each of the election periods is associated with fairly substantial differences compared with the preceding election periods, and all are actually significantly different at the 95 per cent level. Figure \@ref(fig:muelectionall) compares the topics that are significant in the different Howard terms.

To a certain extent the change after the November 2001 election is expected because of the 9/11 terrorist attacks that had only occurred two months earlier, the Bali Bombings that occurred in October 2002, and the dramatic increase in the discussion related to terrorism and conflict over these years. However, the change in 1998 and 2004 is more unexpected. Although the Howard Premiership is the second-longest serving government and commonly thought of as a period of stability because the senior ministers were consistent as well, it might be that it is better to think of the Howard Premiership as a combination of three or four different periods and that the Howard Premiership reinvented itself over this period.

One advantage of our analysis model compared with using the STM approach is that we can create a measure that is equivalent to testing for outliers in a model where the underlying variables were not latent. The results of this reduction in supervision are promising, but suggest the specifics of our process may need further refinement. For instance, our approach appropriately identifies the sitting day that first follows 11 September 2001 and 12 October 2002, which were the dates of the 9/11 Attacks and Bali Bombings respectively. But there are many dates that we would have expected to be identified that were not, and similarly some of the dates that were identified are surprising. When we dig into this we find that some of them, such as 4 February 2009, are associated with significant legislation. However the over-weighting of dates in the first half of our sample in Table \@ref(tab:keyeventsresultstable) highlights further work is needed to improve this approach. For instance, our approach may not be appropriately considering step-changes or it may be that our identification of sitting periods is not appropriate for the entire period.

@@

@RobertsStewartAiroldiRPackage provides more detail about the diagnostic tests that we use, but we briefly discuss each here. Exclusivity is a measure of how specific words are to particular topics. It looks at the proportion that a word makes up of a particular topic compared with the proportion that word makes up of the other topics. As the number of topics increases we usually expect exclusivity to increase because the topics become more particular. Higher values are better. The held-out likelihood as described by @Wallach2009 takes a test/training approach to estimate the probability of held-out documents given the training documents. Higher values are better. The lower bound gives some indication of whether the model may have multiple modes and hence the end result be sensitive to the starting position (@Roberts2016navigating). Residuals analysis, @Taddy2012, compares the theoretical distribution of the variance with the actual distribution. It is a test for overdispersion of the variance, and if it is found then this can suggest that more topics would be appropriate. Semantic coherence is the trade-off for having topics that are more specific and the subsequent risk that the topics become meaningless. @Mimno2011 define a measure of coherence that is based on ratios of single words compared with pairs of words. The idea is that words that should occur in the one document should be more likely to be in a particular topic than ones that do not occur together. For instance, a topic that has 'wine' and 'cheese' as highly rated words would score better on their measure than another that contained 'cheese' and 'mining'. Lower values are better. @RobertsStewartAiroldiRPackage recommend examining the trade-off between exclusivity and semantic coherence. This suggests that the magnitude of improvement reduces from about 80 topics (Figure \@ref(fig:topicsexclusivityvscoherence)).

@@

## Correlated Topic Model  {#correlatedtopicmodelsection}
One of the limitations of LDA is that the model assumes that the presence of one topic is not correlated with the presence of another topic. In reality often topics are related. For instance, in the Hansard context, we may expect topics related to the army to be more commonly found with topics related to the navy, but less commonly with topics related to banking. The goal of the CTM [@BleiLafferty2007] is to account for this correlation between topics, in order to produce more realistic and stable topic distributions over time. The models are very similar, and the key difference is the underlying distributions that are drawn from.

@@

In Figure \@ref(fig:theeightsnighties) we focus on the period of the 1980s and 1990s when there was a great deal of economic change.

@@

In Figure \@ref(fig:economicsgfc) we focus on the financial crisis of 2007-08.

@@

@@

Between 1901 and 2018 30 different people have been prime minister. Four of those---Deakin, Fisher, Menzies and Rudd---returned to the primiership at least once after being replaced, meaning Scott Morrison defines the 36th prime ministerial period.

Three people have died while prime minister: Lyons in 1939, Curtin in 1945, and Holt in 1967. Their respective immediate successors were only prime minister for a short period. As such, we do not consider them when determining the neighbouring prime minister. Specifically, we compare: the first Menzies premiership with Lyons instead of Page; Chifley with Curtin instead of with Forde; and Gorton with Holt instead of with McEwan.

The other prime ministerial period that we do not consider in this paper is the second Rudd premiership This is because it only contained a few sitting days. That is, we compare Tony Abbot with Julia Gillard.

>>>>

The first panel of Figure \@ref(fig:threeexamples) is the second Menzies term. The dashed lines show the elections. While the results here abstract from differences based on topic mix, the shape of the distribution seems to be reasonably similar. Given 80 topics, it may be difficult to see differences in the topic mix manually, and so we would like for the analysis model to be able to suggest whether the mix is changing. The second panel shows the 1983 election, and the subsequent change from Fraser to Hawke on 11 March 1983, identified by a dotted line on that date. The distribution of topics seems less bunched after the change, but it is difficult to tell how different it is. We would like the analysis model to be able to distinguish between the two if they seem different. The third panel shows the first Rudd term. The important aspect to note is that the period after the 2007 election and the Rudd term contain essentially the same Hansard dates. The analysis model needs to be able to deal with this situation.

@@

governments_results %>%
  kable(booktabs = T, caption = "Prime ministers that were significantly different to their predecessor") %>%
  kable_styling(font_size = 10) %>%
  footnote(
    general = "The significance of a prime minister is determined by whether at least one topic was significantly different during this term, compared with the previous one.",
    footnote_as_chunk = T,
    threeparttable = TRUE)

@@

Of the 36 prime ministers over this period, we find that 20 of them are significantly different to the prime minister that preceded them. However, four of these results -- the significance of the Page term, the first Menzies term, the Chifley term and the Abbott term -- are likely due to the short length of either the prime minister or the predecessor and should be ignored, leaving only 16. The earliest of these are the first Fisher term, the Cook term and the third Fisher term, which were different to the second Deakin term, the second Fisher term, and the Cook term, respectively. To a certain extent this is likely due to changes due to World War I.

The second Menzies term, beginning in 1949, is the next government that is significantly different to its predecessor. The other governments that are different are concentrated in the second half of our sample, with three of them being in the past twenty years. Similarly, of the 45 general elections that have been held we find that 14 of them define periods that were significantly different to their immediate predecessor. 1974, 1980, 1990, 1998, 2004, and 2007 stand out as elections where the government did not change, but the model suggests there was considerable change in the topics discussed in parliament.

The second Menzies term was associated with a variety of changes compared with the preceding Chifley Government. Chifley had governed through the end of World War II and the difficult economic times that followed. There was also a large increase in the number of seats in the House of Representatives at the 1949 election. Many new politicians entered parliament, and this changed representation may also have been partly to do with the changed distribution of discussion topics, although further investigation of this is left for future work. The sixteen-year length of the second Menzies term, and better economic conditions over this time make it understandable why parliamentary discussion would have been different. There were six elections within the second Menzies term. Three toward the middle of that government were associated with significant changes in the topics discussed.

Menzies was succeeded  Holt in January 1966. This is an example where there was a change in prime minister without an election, as the next election only happened in November 1966. We find that Holt is different to Menzies. In Figures \@ref(fig:muelectionall) we compare the topics during Menzies' final term with the topics of Holt.

@@

Whitlam's term is especially interesting as we find a difference in the topics after he was first elected in December 1972, compared with his second election win in May 1974. Figure \@ref(fig:muelectionall) compares the topics that are significant before and after the election.

Howard is also interesting because of the significant differences between elections. For instance, each of the election periods is associated with fairly substantial differences compared with the preceding election periods, and all are actually significantly different at the 95 per cent level. Figure \@ref(fig:muelectionall) compares the topics that are significant in the different elections that Howard won.

To a certain extent the change after the November 2001 election is expected because of the 9/11 terrorist attacks that had only occurred two months earlier, the Bali Bombings that occurred in October 2002, and the dramatic increase in the discussion related to terrorism and conflict over these years. However, the change in 1998 and 2004 is more unexpected. Although Howard is the second-longest serving prime minister and commonly thought of as a period of stability because the senior ministers were consistent as well, it might be that it is better to think of Howard as a combination of three or four different periods and that Howard reinvented itself over this period.

One advantage of our analysis model compared with using the STM approach is that we can create a measure that is equivalent to testing for outliers in a model where the underlying variables were not latent. The results of this reduction in supervision are promising, but suggest the specifics of our process may need further refinement. For instance, our approach appropriately identifies the sitting day that first follows 11 September 2001 and 12 October 2002, which were the dates of the 9/11 Attacks and Bali Bombings respectively. But there are many dates that we would have expected to be identified that were not, and similarly some of the dates that were identified are surprising. When we examine this we find that some of them, such as 4 February 2009, are associated with significant legislation. However the over-weighting of dates in the first half of our sample in Table \@ref(tab:keyeventsresultstable) highlights further work is needed to improve this approach. For instance, our approach may not be appropriately considering step-changes or it may be that our identification of sitting periods is not appropriate for the entire period.

@@

@RobertsStewartAiroldiRPackage provides more detail about the diagnostic tests that we use, but we briefly discuss each here. Exclusivity is a measure of how specific words are to particular topics. It looks at the proportion that a word makes up of a particular topic compared with the proportion that word makes up of the other topics. As the number of topics increases we usually expect exclusivity to increase because the topics become more particular. Higher values are better. The held-out likelihood as described by @Wallach2009 takes a test/training approach to estimate the probability of held-out documents given the training documents. Higher values are better. The lower bound gives some indication of whether the model may have multiple modes and hence the end result be sensitive to the starting position (@Roberts2016navigating). Residuals analysis, @Taddy2012, compares the theoretical distribution of the variance with the actual distribution. It is a test for over-dispersion of the variance, and if it is found then this can suggest that more topics would be appropriate. Semantic coherence is the trade-off for having topics that are more specific and the subsequent risk that the topics become meaningless. @Mimno2011 define a measure of coherence that is based on ratios of single words compared with pairs of words. The idea is that words that should occur in the one document should be more likely to be in a particular topic than ones that do not occur together. For instance, a topic that has 'wine' and 'cheese' as highly rated words would score better on their measure than another that contained 'cheese' and 'mining'. Lower values are better. @RobertsStewartAiroldiRPackage recommend examining the trade-off between exclusivity and semantic coherence. This suggests that the magnitude of improvement reduces from about 80 topics (Figure \@ref(fig:topicsexclusivityvscoherence)).

@@

## Correlated Topic Model  {#correlatedtopicmodelsection}
One of the limitations of LDA is that the model assumes that the presence of one topic is not correlated with the presence of another topic. In reality, topics are often related. For instance, in the Hansard context, we may expect topics related to the army to be more commonly found with topics related to the navy, but less commonly with topics related to banking. The goal of the CTM [@BleiLafferty2007] is to account for this correlation between topics, in order to produce more realistic and stable topic distributions over time. The models are very similar, and the key difference is the underlying distributions that are drawn from.

@@

The notable aspect of the response to the 1929 'Black Tuesday' stock market declines was the Premiers' Plan, which appeared newspapers on 11 June 1931. @Copland1934 describes the differences of opinion that occurred in the lead-up to the plan, but Figure \@ref(fig:greatd) highlights how new the Scullin Government was when the economic troubles began.

In Figure \@ref(fig:theeightsnighties) we focus on the 1980s and 1990s when there was a great deal of economic change.

@@

One notable aspect is the difference between the House of Representatives and the Senate that the proportion of economic topics accounts for. The emphasis that the Hawke/Keating government placed on economic issues comes through clearly given the clear change following Hawke's election.

Finally, in Figure \@ref(fig:economicsgfc) we focus on the financial crisis of 2007-08.

@@

It is notable that despite the enormous importance of the financial crisis of 2007-08, topics that are clearly related to economic issues do not dominate the discussion in either house.

@@

Between 1901 and 2018 30 different people have been prime minister. Four of those---Deakin, Fisher, Menzies and Rudd---returned as prime minister at least once after being replaced, meaning Scott Morrison defines the 36th prime ministerial period.

Three people have died while prime minister: Lyons in 1939, Curtin in 1945, and Holt in 1967. Their respective immediate successors were only prime minister for a short period. As such, we do not consider them when determining the neighbouring prime minister. Specifically, we compare: the first Menzies term with Lyons instead of Page; Chifley with Curtin instead of with Forde; and Gorton with Holt instead of with McEwan.

The other prime ministerial period that we do not consider in this paper is the second Rudd term This is because it only contained a few sitting days. That is, we compare Tony Abbot with Julia Gillard.
